---
layout: post
title:  Visualizing R-Ladies's growth! 
date: "2017-04-20 12:11:29 UYT"
published: false
tags: [rstats, r, gganimate, plotly]
description: How to plot animated maps in R, with gganimate and plotly.
---
Here I show how to plot the Twitter users' data retrieved in my previous post in different animated maps, using `gganimate` and `plotly`.

<!--more-->

As I mentioned in my previous post, I recently [came across this article](http://spatial.ly/2017/03/mapping-5000-years-of-city-growth/), and I knew I had to produce a similar map for the [R-Ladies' chapters](http://rladies.org/). I insist that the purple color did its magic with me. So my idea was to plot all the R-Ladies' chapters according to their size, and that's when I thought of using their Twitter followers as a way to estimate it.

The first part of this project was to get all the R-Ladies chapters' Twitter users, and that is [what I did in my previous post]({% post_url 2017-04-20-how-to-fetch-twitter-users-with-r %}){:target="_blank"}. Now I have to plot the maps and animate them. Fun!

# Using Plotly

Let's take a look at the data from my last blog post:

```{r load_data, echo = FALSE, message = FALSE, warning = FALSE}
# You can find everything I use here:
# https://github.com/d4tagirl/R-Ladies-growth-maps

library(knitr)
knitr::opts_chunk$set(dpi = 130, fig.align = 'center', screenshot.force = FALSE, fig.height = 4, fig.cap = "")
options(width = 80, dplyr.width = 150)

rladies <- readRDS('/Users/Daniela/rladies_growing/rladies.rds')
```

```{r  head_users, message = FALSE, warning = FALSE}
library(knitr)

kable(head(rladies), format = "html")
```

I want to produce a map where I can plot each chapter according to its location, with the point's size indicating the amount of Twitter followers. I'm chosing `plotly` now because I love its interactivity, and you can select what you want to see when you hover over the points.

I use the `maps` package to get the world map, using `ggplot2` and `theme_map` for plotting it nicely. 

```{r message = FALSE, warning = FALSE}
library(ggplot2)
library(maps)
library(ggthemes)

world <- ggplot() +
  borders("world", colour = "gray85", fill = "gray80") +
  theme_map() 
```

```{r echo = FALSE}
world
```

Now I plot the chapters. I chose the purple color, obviously!

```{r}
map <- world +
  geom_point(aes(x = lon, y = lat,
                 size = followers),
             data = rladies, colour = 'purple', alpha = .5) +
  scale_size_continuous(range = c(1, 10), breaks = c(250, 500, 750, 1000)) +
  labs(size = 'Followers')
```

```{r echo = FALSE, fig.height = 4}
map
```

The only thing left is to animate it with `plotly`, and this is really easy!!

```{r message = FALSE, warning = FALSE}
library(plotly)

# ggplotly(map, tooltip = c('size'))

```

Now if you hover over the map you can see how many folowers each chapter have. I use the `tooltip` parameter to specify the aesthetics I want `ggplotly` to display in the label when hovering over each point in the map.

But what if you want to see more information about every chapter? There is a very simple way to do this. You only need to add an additional aesthetic parameter to the `ggplot`, and it will pass it to `ggplotly`.

```{r message = FALSE, warning = FALSE, screenshot.force = FALSE}
map <- world +
  geom_point(aes(x = lon, y = lat, 
                 text = paste('city: ', location, "<br />",
                  'followers: ', followers),
                 size = followers,
                 creation = created_at),
             data = rladies, colour = 'purple', alpha = .5) +
  scale_size_continuous(range = c(1, 10), breaks = c(250, 500, 750, 1000)) +
  labs(size = 'Followers')

# ggplotly(map, tooltip = c('text', 'size', 'creation')) 
```

Here I add two new parameters to the aesthetics: the first one is `text`, that allowes me to customize output using `paste()`, and the second one that I invented is `creation`, and it is passed just like that to the tooltip. I couldn't find a way to do the same I did with `text`, with a different parameter... The creators even call it _"the unofficial "text" aesthetic"_, so I don't know if there actually you can do that woth a different parameter.

```{r}
library(ggplot2)
library(plotly)

g <- list(
  showframe = FALSE,
  coastlinecolor = toRGB("grey"),
  showland = TRUE,
  landcolor = toRGB("grey"),
  showcountries = TRUE,
  countrycolor = toRGB("white"),
  countrywidth = 0.2,
  projection = list(type = 'Mercator')
)

p <- plot_geo(rladies,
              marker = list(color = "#B23AEE",
                            opacity = 0.5,
                            line = list(color = "#7A378B",
                                        width = 1.5))) %>%
  add_markers(
    x = ~lon,
    y = ~lat,
    sizes = c(1, 400),
    size = ~followers,
    hoverinfo = "text",
    text = ~paste('city: ', rladies$location, "<br />",
                  'followers: ', rladies$followers)
  ) %>%
  layout(
    geo = g
  )

p

```

```{r}

# #··············
# # gganimate 
# 
# map <- ggplot(world.cities, package = 'maps') +
#   borders('world', colour = 'gray80', fill = 'gray80') +
#   theme_map() +
#   geom_point(aes(x = lon, y = lat, text = paste('city: ', location),
#                  size = followers,
#                  frame = created_at,
#                  cumulative = TRUE),
#              data = rladies, colour = 'purple', alpha = .5) +
#   scale_size_continuous(range = c(1, 10), breaks = c(250, 500, 750, 1000)) + 
#   labs(size = 'Followers')
# 
# library(gganimate)
# 
# # animation::ani.options(ani.width = 1000, ani.height = 600)
# # # gganimate(map, interval = .3)
# # gganimate(map, interval = .3, filename = 'rladies.gif')
# 
# animation::ani.options(ani.width = 750, ani.height = 450)
# gganimate(map, interval = .3, filename = 'rladies.gif')
# 
# #··············
# # gganimate, adding one transparent geom_point frame at the beggining
# 
# # init point to show empty map in the beggining
# ghost_point <- rladies %>%
#   add_row(
#     created_at = format(as.Date('2012-09-01'), format = '%Y-%m-%d'),
#     followers = 0,
#     lon = 0,
#     lat = 0,
#     .before = 1) %>% 
#   slice(1)
# 
# map_ghost <- map + 
#   geom_point(aes(x = lon, y = lat, text = paste('city: ', location), #print init point
#                  size = followers,
#                  frame = created_at,
#                  cumulative = TRUE),
#              data = ghost_point, colour = 'blue', alpha = 0) + 
#   labs(size = 'Followers')
# 
# # animation::ani.options(ani.width = 1000, ani.height = 600)
# # # gganimate(map_ghost, interval = .3)
# # gganimate(map_ghost, interval = .3, filename = 'rladies_ghost.gif')
# 
# animation::ani.options(ani.width = 750, ani.height = 450)
# gganimate(map_ghost, interval = .3, filename = 'rladies_ghost.gif')
# 
# #··············
# # gganimate - with intermediate points!
# 
# library(tibble)
# 
# dates <- as_tibble(seq(as.Date(min(rladies$created_at)), 
#                        as.Date('2017-04-25'), 
#                        by = 'days')) %>% 
#   filter(day(value) %in% c(1, 5, 10, 15, 20, 25))
# 
# rladies_frames <- rladies %>% 
#   nest(-screen_name) %>% 
#   expand(screen_name, date = dates$value) %>%
#   right_join(rladies, by = 'screen_name') %>% 
#   filter(date > created_at) %>% 
#   mutate(date = format(date, format = '%Y-%m-%d'),
#          age_total = as.numeric(age_days, units = 'days'),
#          age_at_date = as.numeric(difftime(date, created_at, unit = 'days'), units = 'days'),
#          est_followers = ((followers - 1) / age_total) * age_at_date)
# 
# # modify init point to show empty map in the beggining
# 
# ghost_point <-  ghost_point %>% 
#   mutate(date = format(created_at, format = '%Y-%m-%d'),
#          est_followers = 0)
# 
# map_frames <- ggplot(world.cities, package = 'maps') +
#   borders('world', colour = 'gray80', fill = 'gray80') +
#   theme_map() +
#   geom_point(aes(x = lon, y = lat, text = paste('city: ', location),
#                  size = est_followers,
#                  frame = date),
#              data = rladies_frames, colour = 'purple', alpha = .5) +
#   geom_point(aes(x = lon, y = lat, text = paste('city: ', location), #print init point
#                  size = est_followers,
#                  frame = date),
#              data = ghost_point, colour = 'blue', alpha = 0) +
#   scale_size_continuous(range = c(1, 10), breaks = c(250, 500, 750, 1000)) + 
#   labs(size = 'Followers')
#              
# # animation::ani.options(ani.width = 1000, ani.height = 600)
# # # gganimate(map_frames, interval = .2)
# # gganimate(map_frames, interval = .2, filename = 'rladies_frames.gif')
# 
# animation::ani.options(ani.width = 750, ani.height = 450)
# gganimate(map_frames, interval = .2, filename = 'rladies_frames.gif')
# 
# #··············
# # gganimate - with intermediate points - leaving some frames before London creation out
# 
# rladies_less_frames <- rladies_frames %>% 
#   filter((day(date) == 1 & month(date) %% 6 == 0) |
#            date >= rladies$created_at[rladies$screen_name == 'RLadiesLondon'])
# 
# map_less_frames <- ggplot(world.cities, package = 'maps') +
#   borders('world', colour = 'gray80', fill = 'gray80') +
#   theme_map() +
#   geom_point(aes(x = lon, y = lat, text = paste('city: ', location),
#                  size = est_followers,
#                  frame = date),
#              data = rladies_less_frames, colour = 'purple', alpha = .5) + 
#   geom_point(aes(x = lon, y = lat, text = paste('city: ', location), #print init point
#                  size = est_followers,
#                  frame = date),
#              data = ghost_point, colour = 'blue', alpha = 0) +
#   scale_size_continuous(range = c(1, 10), breaks = c(250, 500, 750, 1000)) + 
#   labs(size = 'Followers')
# 
# # animation::ani.options(ani.width = 1000, ani.height = 600)
# # # gganimate(map_less_frames, interval = .2)
# # gganimate(map_less_frames, interval = .2, filename = 'rladies_less_frames.gif')
# 
# animation::ani.options(ani.width = 750, ani.height = 450)
# gganimate(map_less_frames, interval = .2, filename = 'rladies_less_frames.gif')
# 
# #··············
# # gganimate -  leaving some frames before London creation out - faster!
# 
# dates <- as_tibble(seq(min(rladies$created_at), 
#                        as.POSIXlt('2017-04-25'), 
#                        by = 'days')) %>% 
#   filter(day(value) %in% c(1, 10, 20))
# 
# rladies_frames <- rladies %>% 
#   nest(-screen_name) %>% 
#   expand(screen_name, date = dates$value) %>%
#   right_join(rladies, by = 'screen_name') %>% 
#   filter(date > created_at) %>% 
#   mutate(date = format(date, format = '%Y-%m-%d'),
#          age_total = as.numeric(age_days, units = 'days'),
#          age_at_date = as.numeric(difftime(date, created_at, unit = 'days'), units = 'days'),
#          est_followers = ((followers - 1) / age_total) * age_at_date)
# 
# rladies_faster <- rladies_frames %>% 
#   filter((day(date) == 1 & month(date) %% 6 == 0) |
#            date >= rladies$created_at[rladies$screen_name == 'RLadiesLondon'])
# 
# map_faster <- ggplot(world.cities, package = 'maps') +
#   borders('world', colour = 'gray80', fill = 'gray80') +
#   theme_map() +
#   geom_point(aes(x = lon, y = lat, text = paste('city: ', location),
#                  size = est_followers,
#                  frame = date),
#              data = rladies_faster, colour = 'purple', alpha = .5) +
#   geom_point(aes(x = lon, y = lat, text = paste('city: ', location), #print init point
#                  size = est_followers,
#                  frame = date),
#              data = ghost_point, colour = 'blue', alpha = 0) +
#   scale_size_continuous(range = c(1, 10), breaks = c(250, 500, 750, 1000)) + 
#   labs(size = 'Followers')
# 
# # animation::ani.options(ani.width = 1000, ani.height = 600)
# # # gganimate(map_less_frames_fast, interval = .2)
# # gganimate(map_less_frames_fast, interval = .2, filename = 'rladies_less_frames_fast.gif')
# 
# animation::ani.options(ani.width = 750, ani.height = 450)
# gganimate(map_faster, interval = .2, filename = 'rladies_faster.gif')
#   


# save.image('RLadies_twitter_growth.RData')
```
