---
layout: post
title:  How to plot animated maps with gganimate
date: "2017-04-24 12:11:29 UYT"
published: true
tags: [rstats, r, gganimate, maps, gif]
description: How to plot an animated map using `gganimate`, and produce a .gif file to share it!
---
Here I show how to plot an animated map using `gganimate`, and produce a .gif file to share it!

<!--more-->

As I mentioned in previous posts, I recently [came across this article](http://spatial.ly/2017/03/mapping-5000-years-of-city-growth/), and I knew I had to produce a similar map for the [R-Ladies' chapters](http://rladies.org/). I insist that the purple color did its magic with me. So my idea was to plot all the R-Ladies' chapters according to their size, and that's when I thought of using their Twitter followers as a way to estimate it.

The first part of this project was to get all the R-Ladies chapters' Twitter users, and that is [what I did in this previous post]({% post_url 2017-04-20-how-to-fetch-twitter-users-with-r %}){:target="_blank"}. Then I plotted them in a map using `ggplot2` and animated the map using the `plotly` package. Despite `plotly::ggplotly` apparently being the easier way, I discovered that it is not the way to go if you want to publish it on a website, as [I examine in this article]({% post_url 2017-04-24-maps-in-plotly-ggplotly-s-huge-maps-or-plot-geo %}){:target="_blank"}. It produces an extremely large HTML, so you should rather use `plotly::plot_geo` instead. 

Now I want to plot a similar map, but animating it using `gganimate`, so I can export it to a .gif file I can easily share. Fun!

# The data

Let's take a look at the data from my last blog post:

```{r load_data, echo = FALSE, message = FALSE, warning = FALSE}
# You can find everything I use here:
# https://github.com/d4tagirl/R-Ladies-growth-maps

library(knitr)

opts_knit$set(animation.fun = hook_ffmpeg_html)

# opts_knit$set(animation.fun = function(x, options, format = "gif") {
#   x = c(knitr:::sans_ext(x), knitr:::file_ext(x))
#   fig.num = options$fig.num
#   format = sub("^[.]", "", format)
#   base = sub(paste0(fig.num, '$'), '', x[1])
#   fig.fname = paste0(sub(paste0(fig.num, "$"), "*", x[1]),
#                      ".", x[2])
#   mov.fname = paste0(sub('-$', '', base), '.', format)
# 
#   # order correctly
#   figs <- Sys.glob(fig.fname)
#   figs <- figs[order(as.numeric(stringr::str_match(figs, paste0("(\\d+)\\.", x[2]))[, 2]))]
# 
#   animation::im.convert(figs, output = mov.fname)
# 
#   sprintf("![%s](%s)", options$label, paste0(opts_knit$get("base.url"), mov.fname))
# })

knitr::opts_chunk$set(dpi = 130, fig.align = 'center', screenshot.force = FALSE, fig.height = 4, fig.cap = "")
options(width = 80, dplyr.width = 150)

rladies <- readRDS('/Users/Daniela/rladies_growing/rladies.rds')
```

```{r  head_users, message = FALSE, warning = FALSE}
library(DT)

datatable(rladies, rownames = FALSE,
          options = list(pageLength = 5))
```

# Plot the map using ggplot2

<br />
I want to produce a map where I can plot each chapter according to its location, with the point's size indicating the amount of Twitter followers. 

I use the `maps` package to get the world map, using the `ggplot2` and `theme_map` ones for plotting it nicely. 

```{r message = FALSE, warning = FALSE}
library(ggplot2)
library(maps)
library(ggthemes)

world <- ggplot() +
  borders("world", colour = "gray85", fill = "gray80") +
  theme_map() 
```

```{r echo = FALSE, message = FALSE, warning = FALSE}
world
```

Now I plot the chapters. I chose the purple color, obviously!

```{r message = FALSE, warning = FALSE}
map <- world +
  geom_point(aes(x = lon, y = lat,
                 size = followers),
             data = rladies, 
             colour = 'purple', alpha = .5) +
  scale_size_continuous(range = c(1, 8), 
                        breaks = c(250, 500, 750, 1000)) +
  labs(size = 'Followers')
```

```{r echo = FALSE, message = FALSE, warning = FALSE}
map
```

With the `range` parameter I control the scale of the points' size.


# Animate the map using gganimate!

Now the only thing left is to animate the map! The core thing here is I want every chapter appearing in the order it was created, to somehow tell a story with the map. Lets start by animating `map`, the `ggplot` object we just created. I have to make a few changes for `gganimate` to work:

- for  `gganimate` to work it needs a `frame` aesthetic: I'll use the `created_at` variable. You set this as a new aesthetic in the `ggplot` and it is ignored by it, but it passes to the `gganimate`;
- I also added the `cumulative = TRUE`, so once the chapter appears on the map, it remains.

```{r}
# 
# # esta anda pero me deja cada frame separado. supuestamente se arregla poniendo fig.show = 'animate'. pruebo en la que viene!
# library(gganimate)
# 
# map <- ggplot() +
#   borders('world', colour = 'gray80', fill = 'gray80') +
#   theme_map() +
#   geom_point(aes(x = lon, y = lat, text = paste('city: ', location),
#                  size = followers,
#                  frame = created_at,
#                  cumulative = TRUE),
#              data = rladies, colour = 'purple', alpha = .5) +
#   scale_size_continuous(range = c(1, 10), breaks = c(250, 500, 750, 1000)) +
#   labs(size = 'Followers')
# 
# animation::ani.options(ani.width = 750, ani.height = 450)
# gganimate(map, interval = .3)
```

```{r fig.show = "animate"}

library(gganimate)

map <- ggplot() +
  borders('world', colour = 'gray80', fill = 'gray80') +
  theme_map() +
  geom_point(aes(x = lon, y = lat, text = paste('city: ', location),
                 size = followers,
                 frame = created_at,
                 cumulative = TRUE),
             data = rladies, colour = 'purple', alpha = .5) +
  scale_size_continuous(range = c(1, 10), breaks = c(250, 500, 750, 1000)) +
  labs(size = 'Followers')

animation::ani.options(ani.width = 750, ani.height = 450)
gganimate(map, interval = .3)
```


```{r fig.show = "animate"}

# library(gganimate)
# 
# map <- ggplot() +
#   borders('world', colour = 'gray80', fill = 'gray80') +
#   theme_map() +
#   geom_point(aes(x = lon, y = lat, text = paste('city: ', location),
#                  size = followers,
#                  frame = created_at,
#                  cumulative = TRUE),
#              data = rladies, colour = 'purple', alpha = .5) +
#   scale_size_continuous(range = c(1, 10), breaks = c(250, 500, 750, 1000)) +
#   labs(size = 'Followers')
# 
# animation::ani.options(ani.width = 750, ani.height = 450)
# gganimate(map, interval = .3)

# gganimate(map, interval = .3, filename = 'rladies.gif')
# 
# animation::ani.options(ani.width = 750, ani.height = 450)
# gganimate(map, interval = .3, filename = 'rladies.gif')

```
 
```{r}
# world <- ggplot() +
#   borders("world", colour = "gray85", fill = "gray80") +
#   theme_map()

# map <- ggplot() +
#   borders('world', colour = 'gray80', fill = 'gray80') +
#   theme_map() +
#   geom_point(aes(x = lon, y = lat, text = paste('city: ', location),
#                  size = followers,
#                  frame = created_at,
#                  cumulative = TRUE),
#              data = rladies, colour = 'purple', alpha = .5) +
#   scale_size_continuous(range = c(1, 10), breaks = c(250, 500, 750, 1000)) + 
#   labs(size = 'Followers')
# 
# library(gganimate)
# 
# # animation::ani.options(ani.width = 1000, ani.height = 600)
# # # gganimate(map, interval = .3)
# # gganimate(map, interval = .3, filename = 'rladies.gif')
# 
# animation::ani.options(ani.width = 750, ani.height = 450)
# gganimate(map, interval = .3, filename = 'rladies.gif')
# 
# #··············
# # gganimate, adding one transparent geom_point frame at the beggining
# 
# # init point to show empty map in the beggining
# ghost_point <- rladies %>%
#   add_row(
#     created_at = format(as.Date('2012-09-01'), format = '%Y-%m-%d'),
#     followers = 0,
#     lon = 0,
#     lat = 0,
#     .before = 1) %>% 
#   slice(1)
# 
# map_ghost <- map + 
#   geom_point(aes(x = lon, y = lat, text = paste('city: ', location), #print init point
#                  size = followers,
#                  frame = created_at,
#                  cumulative = TRUE),
#              data = ghost_point, colour = 'blue', alpha = 0) + 
#   labs(size = 'Followers')
# 
# # animation::ani.options(ani.width = 1000, ani.height = 600)
# # # gganimate(map_ghost, interval = .3)
# # gganimate(map_ghost, interval = .3, filename = 'rladies_ghost.gif')
# 
# animation::ani.options(ani.width = 750, ani.height = 450)
# gganimate(map_ghost, interval = .3, filename = 'rladies_ghost.gif')
# 
# #··············
# # gganimate - with intermediate points!
# 
# library(tibble)
# 
# dates <- as_tibble(seq(as.Date(min(rladies$created_at)), 
#                        as.Date('2017-04-25'), 
#                        by = 'days')) %>% 
#   filter(day(value) %in% c(1, 5, 10, 15, 20, 25))
# 
# rladies_frames <- rladies %>% 
#   nest(-screen_name) %>% 
#   expand(screen_name, date = dates$value) %>%
#   right_join(rladies, by = 'screen_name') %>% 
#   filter(date > created_at) %>% 
#   mutate(date = format(date, format = '%Y-%m-%d'),
#          age_total = as.numeric(age_days, units = 'days'),
#          age_at_date = as.numeric(difftime(date, created_at, unit = 'days'), units = 'days'),
#          est_followers = ((followers - 1) / age_total) * age_at_date)
# 
# # modify init point to show empty map in the beggining
# 
# ghost_point <-  ghost_point %>% 
#   mutate(date = format(created_at, format = '%Y-%m-%d'),
#          est_followers = 0)
# 
# map_frames <- ggplot(world.cities, package = 'maps') +
#   borders('world', colour = 'gray80', fill = 'gray80') +
#   theme_map() +
#   geom_point(aes(x = lon, y = lat, text = paste('city: ', location),
#                  size = est_followers,
#                  frame = date),
#              data = rladies_frames, colour = 'purple', alpha = .5) +
#   geom_point(aes(x = lon, y = lat, text = paste('city: ', location), #print init point
#                  size = est_followers,
#                  frame = date),
#              data = ghost_point, colour = 'blue', alpha = 0) +
#   scale_size_continuous(range = c(1, 10), breaks = c(250, 500, 750, 1000)) + 
#   labs(size = 'Followers')
#              
# # animation::ani.options(ani.width = 1000, ani.height = 600)
# # # gganimate(map_frames, interval = .2)
# # gganimate(map_frames, interval = .2, filename = 'rladies_frames.gif')
# 
# animation::ani.options(ani.width = 750, ani.height = 450)
# gganimate(map_frames, interval = .2, filename = 'rladies_frames.gif')
# 
# #··············
# # gganimate - with intermediate points - leaving some frames before London creation out
# 
# rladies_less_frames <- rladies_frames %>% 
#   filter((day(date) == 1 & month(date) %% 6 == 0) |
#            date >= rladies$created_at[rladies$screen_name == 'RLadiesLondon'])
# 
# map_less_frames <- ggplot(world.cities, package = 'maps') +
#   borders('world', colour = 'gray80', fill = 'gray80') +
#   theme_map() +
#   geom_point(aes(x = lon, y = lat, text = paste('city: ', location),
#                  size = est_followers,
#                  frame = date),
#              data = rladies_less_frames, colour = 'purple', alpha = .5) + 
#   geom_point(aes(x = lon, y = lat, text = paste('city: ', location), #print init point
#                  size = est_followers,
#                  frame = date),
#              data = ghost_point, colour = 'blue', alpha = 0) +
#   scale_size_continuous(range = c(1, 10), breaks = c(250, 500, 750, 1000)) + 
#   labs(size = 'Followers')
# 
# # animation::ani.options(ani.width = 1000, ani.height = 600)
# # # gganimate(map_less_frames, interval = .2)
# # gganimate(map_less_frames, interval = .2, filename = 'rladies_less_frames.gif')
# 
# animation::ani.options(ani.width = 750, ani.height = 450)
# gganimate(map_less_frames, interval = .2, filename = 'rladies_less_frames.gif')
# 
# #··············
# # gganimate -  leaving some frames before London creation out - faster!
# 
# dates <- as_tibble(seq(min(rladies$created_at), 
#                        as.POSIXlt('2017-04-25'), 
#                        by = 'days')) %>% 
#   filter(day(value) %in% c(1, 10, 20))
# 
# rladies_frames <- rladies %>% 
#   nest(-screen_name) %>% 
#   expand(screen_name, date = dates$value) %>%
#   right_join(rladies, by = 'screen_name') %>% 
#   filter(date > created_at) %>% 
#   mutate(date = format(date, format = '%Y-%m-%d'),
#          age_total = as.numeric(age_days, units = 'days'),
#          age_at_date = as.numeric(difftime(date, created_at, unit = 'days'), units = 'days'),
#          est_followers = ((followers - 1) / age_total) * age_at_date)
# 
# rladies_faster <- rladies_frames %>% 
#   filter((day(date) == 1 & month(date) %% 6 == 0) |
#            date >= rladies$created_at[rladies$screen_name == 'RLadiesLondon'])
# 
# map_faster <- ggplot(world.cities, package = 'maps') +
#   borders('world', colour = 'gray80', fill = 'gray80') +
#   theme_map() +
#   geom_point(aes(x = lon, y = lat, text = paste('city: ', location),
#                  size = est_followers,
#                  frame = date),
#              data = rladies_faster, colour = 'purple', alpha = .5) +
#   geom_point(aes(x = lon, y = lat, text = paste('city: ', location), #print init point
#                  size = est_followers,
#                  frame = date),
#              data = ghost_point, colour = 'blue', alpha = 0) +
#   scale_size_continuous(range = c(1, 10), breaks = c(250, 500, 750, 1000)) + 
#   labs(size = 'Followers')
# 
# # animation::ani.options(ani.width = 1000, ani.height = 600)
# # # gganimate(map_less_frames_fast, interval = .2)
# # gganimate(map_less_frames_fast, interval = .2, filename = 'rladies_less_frames_fast.gif')
# 
# animation::ani.options(ani.width = 750, ani.height = 450)
# gganimate(map_faster, interval = .2, filename = 'rladies_faster.gif')
# 
# save.image('RLadies_twitter_growth.RData')
```








<blockquote class="twitter-tweet tw-align-center" data-lang="en"><p lang="en" dir="ltr"><a href="https://twitter.com/hashtag/RLadies?src=hash">#RLadies</a> is growing! A sneak peek to my next blog post ;) <a href="https://twitter.com/hashtag/gganimate?src=hash">#gganimate</a> <a href="https://twitter.com/hashtag/rstats?src=hash">#rstats</a> <a href="https://t.co/rIJ02WzRTx">pic.twitter.com/rIJ02WzRTx</a></p>&mdash; Daniela Vázquez (@d4tagirl) <a href="https://twitter.com/d4tagirl/status/853003269953789952">April 14, 2017</a></blockquote>
<script async src="//platform.twitter.com/widgets.js" charset="utf-8"></script>

