---
layout: post
title:  How to plot animated maps with gganimate
date: "2017-04-24 12:11:29 UYT"
published: true
tags: [rstats, r, gganimate, maps, gif]
description: How to plot an animated map using `gganimate`, and produce a .gif file to share it!
---
Here I show how to plot an animated map using `gganimate`, and produce a .gif file to share it!

<!--more-->

As I mentioned in previous posts, I recently [came across this article](http://spatial.ly/2017/03/mapping-5000-years-of-city-growth/), and I knew I had to produce a similar map for the [R-Ladies' chapters](http://rladies.org/). I insist that the purple color did its magic with me `r emo::ji("stuck_out_tongue_closed_eyes")` . So my idea was to plot all the R-Ladies' chapters according to their size, using their Twitter followers as a way to estimate it, since it's the social media we use by excelence (except for some specific chapters).

The first part of this project was to get all the R-Ladies chapters' Twitter users, and that is [what I did in this previous post]({% post_url 2017-04-20-how-to-fetch-twitter-users-with-r %}){:target="_blank"}. Then I plotted them in a map using `ggplot2` and animated the map using the `plotly` package. Despite `plotly::ggplotly` apparently being the easier way, I discovered that it is not the way to go if you want to publish it on a website, as [I examine in this article]({% post_url 2017-04-26-how-to-deal-with-ggplotly-huge-maps %}){:target="_blank"}. It produces an extremely large HTML, so you should rather use `plotly::plot_geo` instead. 

Now I want to plot a similar map, but animating it using the `gganimate` package, that produces a different kind of animation, and I can export it to a .gif file I can easily share. Fun!

# The data

Let's take a look at the data from my last blog post:

```{r load_data, echo = FALSE, message = FALSE, warning = FALSE}
# You can find everything I use here:
# https://github.com/d4tagirl/R-Ladies-growth-maps

library(knitr)

# This a took from gganimate documentation https://github.com/dgrtwo/gganimate/blob/master/README.Rmd

library(animation)
ani.options(autobrowse = FALSE, interval = 1)

opts_knit$set(animation.fun = function(x, options, format = "gif") {
  x = c(knitr:::sans_ext(x), knitr:::file_ext(x))
  fig.num = options$fig.num
  format = sub("^[.]", "", format)
  base = sub(paste0(fig.num, '$'), '', x[1])
  fig.fname = paste0(sub(paste0(fig.num, "$"), "*", x[1]),
                     ".", x[2])
  mov.fname = paste0(sub('-$', '', base), '.', format)

  # order correctly
  figs <- Sys.glob(fig.fname)
  figs <- figs[order(as.numeric(stringr::str_match(figs, paste0("(\\d+)\\.", x[2]))[, 2]))]

  animation::im.convert(figs, output = mov.fname)
  original_path <- basename(mov.fname)
  file.rename(original_path, mov.fname)
  file.remove(figs)

  sprintf("![%s](%s)", options$label, paste0(opts_knit$get("base.url"), mov.fname))
})

knitr::opts_chunk$set(dpi = 130, fig.align = 'center', screenshot.force = FALSE, fig.height = 4, fig.cap = "")
options(width = 80, dplyr.width = 150)

```

```{r  head_users, message = FALSE, warning = FALSE}
library(readr)
library(dplyr)

url_csv <- 'https://raw.githubusercontent.com/d4tagirl/R-Ladies-growth-maps/master/rladies.csv'
rladies <- read_csv(url(url_csv)) %>% 
  select(-1) %>% 
  slice(1:3)

library(DT)

datatable(rladies, rownames = FALSE,
          options = list(pageLength = 5))
```

<br />
# Plotting the map using ggplot2

I want to produce a map where I can plot each chapter according to its location, with the point's size indicating the amount of Twitter followers. 

I use the `maps` package to get the world map, using the `ggplot2::ggplot` and `ggthemes::theme_map` functions for plotting it nicely. 

```{r message = FALSE, warning = FALSE}
library(ggplot2)
library(maps)
library(ggthemes)

world <- ggplot() +
  borders("world", colour = "gray85", fill = "gray80") +
  theme_map() 
```

```{r echo = FALSE, message = FALSE, warning = FALSE}
world
```

Now I plot the chapters. I chose the purple color, obviously!

```{r message = FALSE, warning = FALSE}
map <- world +
  geom_point(aes(x = lon, y = lat,
                 size = followers),
             data = rladies, 
             colour = 'purple', alpha = .5) +
  scale_size_continuous(range = c(1, 8), 
                        breaks = c(250, 500, 750, 1000)) +
  labs(size = 'Followers')
```

```{r echo = FALSE, message = FALSE, warning = FALSE}
map
```

The `range` parameter is what controls the scale of the points' size.


# Animating the map using gganimate!

Now the only thing left is to animate the map! The core thing here is that I want every chapter appearing in the order it was created, to somehow tell a story with the map. Lets start by animating `map`: the `ggplot` object we just created. I have to make a few changes for `gganimate` to work:

- `gganimate` requires a `frame` aesthetic: I'll use the `created_at` variable. You set this as a new aesthetic in `ggplot` which is ignored by it (as shown in the warning message), but `gganimate` recognizes and uses it;

- I also add the `cumulative = TRUE`, an additional aesthetic  (same comment about `ggplot` ignoring it), so once the chapter appears on the map, it keeps showing in all following frames.


```{r map_ani, fig.show = "animate", fig.align = "center", message = FALSE}
map <- world +
  geom_point(aes(x = lon, y = lat,
                 size = followers,
                 frame = created_at,
                 cumulative = TRUE),
             data = rladies, colour = 'purple', alpha = .5) +
  scale_size_continuous(range = c(1, 10), breaks = c(250, 500, 750, 1000)) +
  labs(size = 'Followers')

library(gganimate)

animation::ani.options(ani.width = 750, ani.height = 450)
gganimate(map, interval = .2)
```

This animation is so nice! It shows how R-Ladies is spreading all over the globe, giving also some idea of how it is growing.

# Customizing the gganimate map

Now it is when I start to worry about some details... The first thing was that I showed the animation to a friend, and he told me that there was no frame with an empty map at the beggining. And this was a very good idea!

The first thing to do was to generate a dataframe with some random data (this data will not be visible), with the same structure than the original dataframe, as follows.

```{r}

ghost_point <- rladies %>%
  add_row(
    created_at = format(as.Date('2012-09-01'), format = '%Y-%m-%d'),
    followers = 0,
    lon = 0,
    lat = 0,
    .before = 1) %>%
  slice(1)

```

Then I add an extra layer to the `ggplot` object `map`, but with the `alpha` parameter set on 0, so the point will not show in the plot.

```{r ghost_ani}
map_ghost <- map +
  geom_point(aes(x = lon, y = lat,
                 size = followers,
                 frame = created_at,
                 cumulative = TRUE),
             data = ghost_point, colour = 'blue', alpha = 0) +
  labs(size = 'Followers')

animation::ani.options(ani.width = 750, ani.height = 450)
gganimate(map_ghost, interval = .2)
```

```{r}
# # animation::ani.options(ani.width = 1000, ani.height = 600)
# # # gganimate(map_ghost, interval = .3)
# # gganimate(map_ghost, interval = .3, filename = 'rladies_ghost.gif')
# 
# animation::ani.options(ani.width = 750, ani.height = 450)
# gganimate(map_ghost, interval = .3, filename = 'rladies_ghost.gif')
# 
# #··············
# # gganimate - with intermediate points!
# 
# library(tibble)
# 
# dates <- as_tibble(seq(as.Date(min(rladies$created_at)), 
#                        as.Date('2017-04-25'), 
#                        by = 'days')) %>% 
#   filter(day(value) %in% c(1, 5, 10, 15, 20, 25))
# 
# rladies_frames <- rladies %>% 
#   nest(-screen_name) %>% 
#   expand(screen_name, date = dates$value) %>%
#   right_join(rladies, by = 'screen_name') %>% 
#   filter(date > created_at) %>% 
#   mutate(date = format(date, format = '%Y-%m-%d'),
#          age_total = as.numeric(age_days, units = 'days'),
#          age_at_date = as.numeric(difftime(date, created_at, unit = 'days'), units = 'days'),
#          est_followers = ((followers - 1) / age_total) * age_at_date)
# 
# # modify init point to show empty map in the beggining
# 
# ghost_point <-  ghost_point %>% 
#   mutate(date = format(created_at, format = '%Y-%m-%d'),
#          est_followers = 0)
# 
# map_frames <- ggplot(world.cities, package = 'maps') +
#   borders('world', colour = 'gray80', fill = 'gray80') +
#   theme_map() +
#   geom_point(aes(x = lon, y = lat, text = paste('city: ', location),
#                  size = est_followers,
#                  frame = date),
#              data = rladies_frames, colour = 'purple', alpha = .5) +
#   geom_point(aes(x = lon, y = lat, text = paste('city: ', location), #print init point
#                  size = est_followers,
#                  frame = date),
#              data = ghost_point, colour = 'blue', alpha = 0) +
#   scale_size_continuous(range = c(1, 10), breaks = c(250, 500, 750, 1000)) + 
#   labs(size = 'Followers')
#              
# # animation::ani.options(ani.width = 1000, ani.height = 600)
# # # gganimate(map_frames, interval = .2)
# # gganimate(map_frames, interval = .2, filename = 'rladies_frames.gif')
# 
# animation::ani.options(ani.width = 750, ani.height = 450)
# gganimate(map_frames, interval = .2, filename = 'rladies_frames.gif')
# 
# #··············
# # gganimate - with intermediate points - leaving some frames before London creation out
# 
# rladies_less_frames <- rladies_frames %>% 
#   filter((day(date) == 1 & month(date) %% 6 == 0) |
#            date >= rladies$created_at[rladies$screen_name == 'RLadiesLondon'])
# 
# map_less_frames <- ggplot(world.cities, package = 'maps') +
#   borders('world', colour = 'gray80', fill = 'gray80') +
#   theme_map() +
#   geom_point(aes(x = lon, y = lat, text = paste('city: ', location),
#                  size = est_followers,
#                  frame = date),
#              data = rladies_less_frames, colour = 'purple', alpha = .5) + 
#   geom_point(aes(x = lon, y = lat, text = paste('city: ', location), #print init point
#                  size = est_followers,
#                  frame = date),
#              data = ghost_point, colour = 'blue', alpha = 0) +
#   scale_size_continuous(range = c(1, 10), breaks = c(250, 500, 750, 1000)) + 
#   labs(size = 'Followers')
# 
# # animation::ani.options(ani.width = 1000, ani.height = 600)
# # # gganimate(map_less_frames, interval = .2)
# # gganimate(map_less_frames, interval = .2, filename = 'rladies_less_frames.gif')
# 
# animation::ani.options(ani.width = 750, ani.height = 450)
# gganimate(map_less_frames, interval = .2, filename = 'rladies_less_frames.gif')
# 
# #··············
# # gganimate -  leaving some frames before London creation out - faster!
# 
# dates <- as_tibble(seq(min(rladies$created_at), 
#                        as.POSIXlt('2017-04-25'), 
#                        by = 'days')) %>% 
#   filter(day(value) %in% c(1, 10, 20))
# 
# rladies_frames <- rladies %>% 
#   nest(-screen_name) %>% 
#   expand(screen_name, date = dates$value) %>%
#   right_join(rladies, by = 'screen_name') %>% 
#   filter(date > created_at) %>% 
#   mutate(date = format(date, format = '%Y-%m-%d'),
#          age_total = as.numeric(age_days, units = 'days'),
#          age_at_date = as.numeric(difftime(date, created_at, unit = 'days'), units = 'days'),
#          est_followers = ((followers - 1) / age_total) * age_at_date)
# 
# rladies_faster <- rladies_frames %>% 
#   filter((day(date) == 1 & month(date) %% 6 == 0) |
#            date >= rladies$created_at[rladies$screen_name == 'RLadiesLondon'])
# 
# map_faster <- ggplot(world.cities, package = 'maps') +
#   borders('world', colour = 'gray80', fill = 'gray80') +
#   theme_map() +
#   geom_point(aes(x = lon, y = lat, text = paste('city: ', location),
#                  size = est_followers,
#                  frame = date),
#              data = rladies_faster, colour = 'purple', alpha = .5) +
#   geom_point(aes(x = lon, y = lat, text = paste('city: ', location), #print init point
#                  size = est_followers,
#                  frame = date),
#              data = ghost_point, colour = 'blue', alpha = 0) +
#   scale_size_continuous(range = c(1, 10), breaks = c(250, 500, 750, 1000)) + 
#   labs(size = 'Followers')
# 
# # animation::ani.options(ani.width = 1000, ani.height = 600)
# # # gganimate(map_less_frames_fast, interval = .2)
# # gganimate(map_less_frames_fast, interval = .2, filename = 'rladies_less_frames_fast.gif')
# 
# animation::ani.options(ani.width = 750, ani.height = 450)
# gganimate(map_faster, interval = .2, filename = 'rladies_faster.gif')
# 
# save.image('RLadies_twitter_growth.RData')
```








<blockquote class="twitter-tweet tw-align-center" data-lang="en"><p lang="en" dir="ltr"><a href="https://twitter.com/hashtag/RLadies?src=hash">#RLadies</a> is growing! A sneak peek to my next blog post ;) <a href="https://twitter.com/hashtag/gganimate?src=hash">#gganimate</a> <a href="https://twitter.com/hashtag/rstats?src=hash">#rstats</a> <a href="https://t.co/rIJ02WzRTx">pic.twitter.com/rIJ02WzRTx</a></p>&mdash; Daniela Vázquez (@d4tagirl) <a href="https://twitter.com/d4tagirl/status/853003269953789952">April 14, 2017</a></blockquote>
<script async src="//platform.twitter.com/widgets.js" charset="utf-8"></script>

