---
layout: post
title:  How to deal with ggplotly huge maps
date: "2017-04-24 12:11:29 UYT"
published: true
tags: [rstats, r, plotly, ggplotly, maps, plot_geo]
description: How to produce animated maps in R with plotly, using ggplotly and plot_geo. While ggplotly is easier to use, it produces huge maps (using HTML widgets). Using plot_geo instead is incredibly more efficient.
---
How I had  to go from `gglotly` to `plotly::plot_geo` to reduce the size of the HTML output for posting animated maps on my blog.

<!--more-->
I've been experimenting with animating maps these past few days and I finally was able to produce some fancy maps using `plotly` and `gganimate` ([you can see my repo here](https://github.com/d4tagirl/R-Ladies-growth-maps)). Everything was working perfectly on Rstudio, until I wanted to share some kind of tutorial on my blog, and things got complicated. Originally I animated some maps using `ggplotly` and it generated a 3.3 MB HTML widget, and that's an issue when you want to include it on a website! So I tried a different approach: using `plotly::plot_geo` I was able to produce a very similar plot, but generating a way smaller HTML widget.

# The data

I'm using the [data produced in this previous post]({% post_url 2017-04-20-how-to-fetch-twitter-users-with-r %}){:target="_blank"}. Let's take a look at it to see what we are dealing with.

```{r load_data, echo = FALSE, message = FALSE, warning = FALSE}
# You can find everything I use here:
# https://github.com/d4tagirl/R-Ladies-growth-maps

library(knitr)
knitr::opts_chunk$set(dpi = 130, fig.align = 'center', screenshot.force = FALSE, fig.height = 4, fig.cap = "")
options(width = 80, dplyr.width = 150)

rladies <- readRDS(gzcon(url('https://github.com/d4tagirl/R-Ladies-growth-maps/raw/master/rladies.rds')))

library(plotly)
```

```{r  head_users, message = FALSE, warning = FALSE}
library(knitr)

kable(head(rladies), format = "html")
```

I wanted to produce a map using `plotly` where I could plot each chapter according to its location, with each point's size indicating its Twitter followers. 

# Using ggplotly

The first thing I did was generate the map using `ggplot2`, because `plotly` integrates easily with it. And I'm so comfortable using the `tidyverse`, that it was the natural thing for me to do. 

As I wanted to include in the `plotly` labels variables not in the `ggplot`'s aesthetics, I added the extra `text` parameter for `ggplotly` to include them. `ggplot` just ignores it.

```{r message = FALSE, warning = FALSE}
library(ggplot2)
library(maps)
library(ggthemes)

world <- ggplot() +
  borders("world", colour = "gray85", fill = "gray80") +
  theme_map()

map <- world +
  geom_point(aes(x = lon, y = lat,
                 text = paste('city: ', location,
                              '<br /> created : ', created_at),
                 size = followers),
             data = rladies, colour = 'purple', alpha = .5) +
  scale_size_continuous(range = c(1, 8), breaks = c(250, 500, 750, 1000)) +
  labs(size = 'Followers')
```

```{r echo = FALSE, message = FALSE, warning = FALSE}
map
```

Then I animated it using `ggplotly`, with the following code:

```{r huge_map, eval = FALSE}
library(plotly)

ggplotly(map, tooltip = c('text', 'size'))
```

If you want to see the animation, [you can check it here]({% post_url 2017-04-24-huge-ggplotly-map %}){:target="_blank"}, but it will take a while! It is a 3.3 MB page!

It is actually a pretty nice animation, but it takes for ever to load the HTML! And that is why I checked how the `plotly` people make this kind of animations from scratch, and I gave it a try.

It is actually not difficult either! It was not that straightforward for me how to find the chart references for customize the maps (probably I was not searching properly :-/), so [here is the link](https://plot.ly/r/reference/), and for the layout in particular [here it is this other link](https://plot.ly/r/reference/#layout-geo/) just in case you encounter the same difficulties as I did.

```{r}
g <- list(showframe = FALSE,
          coastlinecolor = toRGB("white"),
          showland = TRUE,
          landcolor = toRGB("gray80"),
          showcountries = TRUE,
          countrycolor = toRGB("white"),
          countrywidth = 0.2,
          projection = list(type = 'Mercator'))

plot_geo(rladies,
         marker = list(color = toRGB("purple"),
                       opacity = 0.5,
                       line = list(color = toRGB("purple"),
                                   width = 1.5))) %>%
  add_markers(x = ~lon,
              y = ~lat,
              sizes = c(1, 450),
              size = ~followers,
              hoverinfo = "text",
              text = ~paste('city: ', location,
                            '<br /> created : ', created_at,
                            '<br /> followers: ', followers)) %>%
  layout(geo = g)
```

This code produces a 16.2 KB HTML, so there I had a 99.5% reduction :)

If you had a different experience, or have any comments about this, please comment below or [mention me on Twitter](https://twitter.com/intent/tweet?user_id=114258616).
