---
layout: post
title:  'Does "the twitter ratio" apply to the #rstats community?'
date: "2018-01-28 12:11:29 UYT"
published: true
tags: [rstats, rtweet, ggtern, purrr, the ratio]
description: "Analysis of the #rstats twitter ratio and visualization of replies, RTs and likes in a pyramid using ggtern, and some particular conversations using interactive graphs."
---

I analize the \#rstats on Twitter to check if "The Ratio" actually makes sense for this community. What is "The Ratio" you ask? There is this theory that if a tweet has more replies than likes and retweets, that implies that people are not happy about it.

<!--more-->

```{r echo = FALSE, message = FALSE, warning = FALSE}
# you can find everything I use here:
# https://github.com/d4tagirl/The-Ratio-rstats-hashtag
library(emo)
library(dplyr)
library(knitr)
knitr::opts_chunk$set(fig.align = 'center', screenshot.force = FALSE, fig.cap = "",
                      dpi = 120)
options(width = 80, dplyr.width = 150)
```

Not long ago I came across [a FiveThirtyEight post called "The Worst Tweeter In Politics Isn’t Trump"](https://fivethirtyeight.com/features/the-worst-tweeter-in-politics-isnt-trump/). Well, it was a long time ago actually, but this project was laying in my computer for a while ¯\\_(ツ)_/¯. They gathered some resources from the media discussing that tweets leading to more replies than likes and retweets are the ones that make the community angry. This phenomenon is known as "The Ratio" as Luke O’Neil wrote recently [in Esquire](http://www.esquire.com/news-politics/news/a54440/twitter-ratio-reply/).

FiveThirtyEight used a ternary plot to illustrate the proportion of replies, retweets and likes of every Trump tweet. In this post I'm going to plot tweets with the rstats hashtag, suspecting that the ones that have a higher ratio of replies might be an exception to this rule since conversations tend to be pretty friendly in this community. But let's find out!

# Retrieving the data

I'm becoming more and more fan of the `rtweet` package, built and maintained by [Michael W. Kearney](https://twitter.com/kearneymw?lang=en). It's _the_ way to go when you want to interact with Twitter's API using R. I fetch some tweets with `#rstats` to analize!

```{r eval=FALSE}
library(rtweet)
library(dplyr)
tweets_rstats <- search_tweets(q = "#rstats",
                                include_rts = FALSE,
                                n = 300)

tweets_rstats <- tweets_rstats %>%
  distinct()
```

```{r message=FALSE, warning=FALSE, echo=FALSE, screenshot.force = FALSE }
library(rtweet)
library(dplyr)

url_tweets_rstats <- 'https://github.com/d4tagirl/The-Ratio-rstats-hashtag/raw/master/tweets_rstats_20180127'
tweets_rstats <- readRDS(url(url_tweets_rstats))

tweets_rstats <- tweets_rstats %>%
  distinct()

library(DT)
datatable(tweets_rstats[, c(4:5)], 
          options = list(pageLength = 3)) %>% 
  formatStyle(c(1:3), `font-size` = '15px')
```
<br/>
I keep only the original tweets with at least two likes because I want to keep _relevant_ tweets. This is probably too arbitrary and surely can be improved, but here I go.  

```{r message=FALSE, warning=FALSE, screenshot.force = FALSE}
orig_tweets <- tweets_rstats %>% 
  filter(is.na(reply_to_status_id),
         favorite_count > 1) %>%    
  select(status_id, screen_name, text, favorite_count, retweet_count) %>%
  distinct()
```

```{r echo = FALSE}
datatable(orig_tweets[, c(1:3)], 
          options = list(pageLength = 3)) %>% 
  formatStyle(c(1:3), `font-size` = '15px')
```

<br/>
I already have the number of retweets and the number of likes of each original tweet, but not the number of replies. To build the ternary plot (or pyramid as I prefer to call it) I need the number of replies as well. As the API doesn't have a direct method to do this, so I have to do it by hand.




Here comes the `purrr` part. The `purrr` package is receiving a lot of love this year: there is a group sharing [the _#purrrResolution_](https://twitter.com/hashtag/purrrresolution?src=hash&lang=en), courtesy of [Isabella Ghement](https://twitter.com/IsabellaGhement/status/949349863853977600) that you can join:

<blockquote class="twitter-tweet tw-align-center" data-lang="en"><p lang="en" dir="ltr">Just sent out the first group e-mail concerning the <a href="https://twitter.com/hashtag/purrrResolution?src=hash&amp;ref_src=twsrc%5Etfw">#purrrResolution</a> <a href="https://twitter.com/hashtag/rstats?src=hash&amp;ref_src=twsrc%5Etfw">#rstats</a> <a href="https://twitter.com/hashtag/purrr?src=hash&amp;ref_src=twsrc%5Etfw">#purrr</a> -  if you haven&#39;t received it, it means you are not yet on the list.  To join the list, you can e-mail me (isabella@ghement.ca).  Keep on purrring!</p>&mdash; Isabella R. Ghement (@IsabellaGhement) <a href="https://twitter.com/IsabellaGhement/status/949349863853977600?ref_src=twsrc%5Etfw">January 5, 2018</a></blockquote>
<script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

And [Colin Fay](https://twitter.com/_ColinFay) created [the Twitter collection _\#RStats — Your daily dose of \#purrr_](https://twitter.com/i/moments/950031718521671680) with great tips!

I collect all the mentions to all `screen_names` in the `orig_tweets` dataframe. I use `distinct(screen_names)` because I don't want to call the API more than once for every `screen_name`.

```{r eval=FALSE}
library(purrr)
library(tidyr)

orig_tweets_mentions <- orig_tweets %>%
  distinct(screen_name) %>%           
  mutate(query = paste0("@", screen_name, " OR ", "to:", screen_name, " OR ", screen_name)) %>%
  mutate(tweets = pmap(list(q = .$query,
                            n = 1000,
                            retryonratelimit = TRUE),
                       rtweet::search_tweets)) %>%
  select(tweets) %>%
  unnest()
```

Here I'm joining the conversation by using the `pmap` function to fetch all the mentions to all the `screen_name`s in the `orig_tweets` dataframe. The API only returns tweets from the last 6 to 10 days, but it should suffice. As [Lucy](https://twitter.com/lucystats?lang=en) pointed out in [her post about Twitter Trees](http://livefreeordichotomize.com/2017/07/24/twitter-trees/), querying the API using only `to: screen_name` misses some tweets, so I took her recommendation of including `@screen_name` and `OR screen_name`. You will notice that I took a lot of ideas from her blog post, which I highly recommend if you like to work with Twitter conversations.

I need to apply the `rtweet::search_tweets` function to each `screen_name` in the `orig_tweets` dataframe, passing more than one argument to the function: `pmap` is the answer! You can pass a list of arguments to the `pmap` function for it to pass them on to the `search_tweets` one. In this case I pass `q`: the query, `n`: the number of tweets I want, and `retryonratelimit`: set to TRUE for it to wait and retry when rate limited.

This is what I get:

```{r message=FALSE, warning=FALSE, screenshot.force = FALSE, echo=FALSE}
url_orig_tweets_mentions <- 'https://github.com/d4tagirl/The-Ratio-rstats-hashtag/raw/master/orig_tweets_mentions_20180127'
orig_tweets_mentions <- readRDS(url(url_orig_tweets_mentions))

orig_tweets_mentions <- distinct(orig_tweets_mentions) 
```

```{r echo = FALSE, warning=FALSE}
datatable(orig_tweets_mentions[c(1:100), c(4:5)], 
          options = list(pageLength = 3)) %>% 
  formatStyle(c(1:3), `font-size` = '15px')
```
<br/>

# Getting the chain of replies

Then I'll use another resource from [Lucy's **genius** post to get the chain of replies from a tweet](http://livefreeordichotomize.com/2017/07/24/twitter-trees/). I create a function that takes the `status_id` as input, and returns all the replies and the replies of that replies and so on. Again I use `purrr` to apply the function to all the `status_id`s from the `orig_tweets` dataframe, but this time I need to pass only one argument to the function, so I use `map` instead of `pmap`.

```{r}
get_replies_chain <- function(id) {
  diff <- 1
  while (diff != 0) {
    id_next <- orig_tweets_mentions %>%
      filter(reply_to_status_id %in% id) %>%
      pull(status_id)
    id_new <- unique(c(id, id_next))
    diff <- length(id_new) - length(id)
    id <- id_new
  }
  orig_tweets_mentions %>% 
    filter(reply_to_status_id %in% id)
}

replies <- orig_tweets %>%
  mutate(replies = purrr::map(.$status_id,
                       get_replies_chain)) %>%
  tidyr::unnest(replies) %>%
  select(status_id, screen_name, text, status_id_reply = status_id1) %>%
  distinct()
```

```{r echo = FALSE}
datatable(replies, 
          options = list(pageLength = 3)) %>% 
  formatStyle(c(1:3), `font-size` = '15px')
```
<br/>
The reason I needed the replies is to count them, because in the original tweet I only have the number of retweets and the number of likes, but not the number of replies. The last step is to count the replies!

```{r}
replies_count <- replies %>%
  group_by(status_id) %>%
  summarise(reply_count = n()) %>%
  ungroup
```

# Building the pyramid

I build a dataframe with the variables I need for the plot. I create the `ratio` variables that is **_The Ratio_**, the proportion of `replies` to `replies + faves` (or likes as we call them now), that is our variable of interest.

```{r}
tweets_tern <- orig_tweets %>%
  left_join(replies_count, by = "status_id") %>%
  mutate(reply_count = coalesce(reply_count, 0L),
         ratio = reply_count / (reply_count + favorite_count)) %>%
  select(screen_name, status_id, 
         replies = reply_count, 
         RTs = retweet_count, 
         faves = favorite_count,
         ratio)
```

```{r echo = FALSE}
datatable(tweets_tern, 
          options = list(pageLength = 3)) %>% 
  formatStyle(c(1:3), `font-size` = '15px')
```
<br/>
I build a different dataframe containing the mean of `replies`, `RTs` and `faves`, for reference.
```{r}
tweets_tern_mean <- tweets_tern %>%
  summarize(mean_replies = mean(replies),
            mean_rt      = mean(RTs),
            mean_fave    = mean(faves))
```

For building the pyramid I use the [ggtern package](http://www.ggtern.com/), an extension to the `ggplot2` package specifically for the plotting of ternary diagrams. I have to use the `lines` dataframe for better visualization, and I plot the mean (`tweets_tern_mean`) bigger.

```{r message=FALSE, warning=FALSE}
library(ggtern)
library(viridis)

lines <- data.frame(x = c(1, 0, 0),
                    y = c(0, 1, 0),
                    z = c(0, 0, 1),
                    xend = c(0, 1, 1),
                    yend = c(1, 0, 1),
                    zend = c(1, 1, 0))

pyramid <- ggtern(data = tweets_tern, aes(x = replies, y = RTs, z = faves)) +
  geom_mask() +
  geom_point(col = ifelse(tweets_tern$replies/(tweets_tern$replies + tweets_tern$faves) > .5, 
                                viridis(5)[3], viridis(5)[1]),
             alpha = 0.7) +
  geom_point(data = tweets_tern_mean,
             aes(mean_replies, mean_rt, mean_fave),
             color = viridis(5)[1], alpha = 0.8, size = 5) +
  theme_classic() +
  theme(legend.position = "none") +
  geom_segment(data = lines,
               aes(x, y, z,
                   xend = xend, yend = yend, zend = zend),
               color = "grey",
               size = .2) +
  theme_showarrows()

pyramid
```

I would have loved to present this ternary plot in an interactive fashion, with a tooltip to see every tweet, but I couldn't find a direct way to do it, so maybe next time I'll figure this out! In the meantime, I decided I could try a different approach to look at these tweets `r emo::ji('sunglasses')`

Apparently there are `r nrow(tweets_tern %>% select(ratio) %>% filter(ratio > .5))` out of the `r nrow(orig_tweets)` with more replies than likes, let's explore them!

# Building Twitter Trees _à la Lucy_ `r emo::ji('dancing_women')`

As I said, I'm a fan of Lucy's post, so I'll plot the conversations that falls into _The Ratio_ rule in the form of a graph, to explore them as she did (with a few tweaks). Also, I'm a bit obsessed with graphs now that I took [James Curley](https://twitter.com/jalapic?ref_src=twsrc%5Egoogle%7Ctwcamp%5Eserp%7Ctwgr%5Eauthorthe)'s ["Network Analysis in R" DataCamp course](https://www.datacamp.com/courses/network-analysis-in-r), which is great! I use `igraph`, `ggraph` and `ggiraph` packages, the last one to make the graphs interactive .

I select the tweets first.

```{r}
tweets_tern_prop <- tweets_tern %>%
  arrange(desc(ratio)) %>%
  filter(ratio > .5)
```

```{r echo = FALSE}
datatable(tweets_tern_prop) %>% 
  formatStyle(c(1:3), `font-size` = '15px')
```
<br/>
then I loop over them to plot each twitter tree. Note that this is an interactive plot, the seed tweet is in a different color, and the size of the point is relative to the likes count.

```{r message=FALSE, warning=FALSE}

library("ggraph")
library("igraph")
library("ggiraph")

set.seed(52)

graphs <- list()

for (i in 1:nrow(tweets_tern_prop)) {
  replies_1 <- get_replies_chain(tweets_tern_prop[i,]$status_id) %>%
    distinct(screen_name, text, status_id, reply_to_status_id, favorite_count)

  tweet_0 <- orig_tweets %>%
    filter(status_id == tweets_tern_prop[i,]$status_id) %>%
    select(screen_name, text, favorite_count)

  from_text <- replies_1 %>%
    select(reply_to_status_id) %>%
    left_join(replies_1, c("reply_to_status_id" = "status_id")) %>%
    select(screen_name, text, favorite_count) %>%
    mutate(favorite_count = coalesce(favorite_count, 0L))

  tweet_0 <- paste0(tweet_0$screen_name, ": ", tweet_0$text, "\nLikes: ", tweet_0$favorite_count)
  
  to_text <- paste0(replies_1$screen_name, ": ", replies_1$text, "\nLikes: ", replies_1$favorite_count)
  to_text <- gsub("'", "`", to_text)
  from_text <- paste0(from_text$screen_name, ": ", from_text$text, "\nLikes: ", from_text$favorite_count)
  from_text <- gsub("'", "`", from_text)

  edges <- tibble::tibble(from = from_text,
                          to = to_text) %>%
    mutate(from = ifelse(from == "NA: NA\nLikes: 0",
                         tweet_0,
                         from))


  graph <- graph_from_data_frame(edges, directed = TRUE)
  V(graph)$tooltip <- V(graph)$name
  V(graph)$tooltip <- gsub("'", "`", V(graph)$tooltip)
  
  library(stringr)
  V(graph)$size <- str_extract(V(graph)$name, "[0-9*]$")

  p <- ggraph(graph, layout = "nicely") +
    geom_edge_link(edge_colour = viridis(5)[3]) +
    geom_point_interactive(aes(x, y,
                               tooltip = tooltip,
                               size = size),
                           color = ifelse(V(graph)$name == tweet_0, viridis(5)[2], viridis(5)[1]),
                           alpha = 0.8) +
                           # , size = 4)
    scale_size_discrete(range = c(3,12)) +
    theme_void() +
    theme(legend.position = "none")
  
  graphs[[i]] <- ggiraph(code = print(p),
                     width_svg = 10,
                     zoom_max = 4)
  }

```

Here are the tweets for us to take a proper look `r emo::ji('tada')`

## First tweet

<blockquote class="twitter-tweet tw-align-center" data-lang="en"><p lang="en" dir="ltr">List columns in the <a href="https://twitter.com/hashtag/tidyverse?src=hash&amp;ref_src=twsrc%5Etfw">#tidyverse</a> is like data.table syntax for me. No matter how many times I use it I feel like I&#39;m figuring out the syntax for the first time. It just doesn&#39;t stick in my brain. <a href="https://twitter.com/hashtag/rstats?src=hash&amp;ref_src=twsrc%5Etfw">#rstats</a></p>&mdash; Joran Elias (@joranelias) <a href="https://twitter.com/joranelias/status/957022615918428160?ref_src=twsrc%5Etfw">January 26, 2018</a></blockquote>
<script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

```{r message=FALSE, warning=FALSE, echo = FALSE}
graphs[[1]]
```

## Second tweet

<blockquote class="twitter-tweet tw-align-center" data-lang="en"><p lang="en" dir="ltr">Let&#39;s play a guessing game. <br><br>What type of data is in this <a href="https://twitter.com/hashtag/Rstats?src=hash&amp;ref_src=twsrc%5Etfw">#Rstats</a> 3d histogram? <a href="https://t.co/y2uJwAmTKP">pic.twitter.com/y2uJwAmTKP</a></p>&mdash; Lynn Mazzoleni (@LynnMazzoleni) <a href="https://twitter.com/LynnMazzoleni/status/957025305847369729?ref_src=twsrc%5Etfw">January 26, 2018</a></blockquote>
<script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>
```{r message=FALSE, warning=FALSE, echo = FALSE}
graphs[[2]]
```
## Third tweet

<blockquote class="twitter-tweet tw-align-center" data-lang="en"><p lang="en" dir="ltr"><a href="https://twitter.com/hashtag/Rstats?src=hash&amp;ref_src=twsrc%5Etfw">#Rstats</a> community: Any recommendations for introductory <a href="https://twitter.com/rstudio?ref_src=twsrc%5Etfw">@rstudio</a> videos for undergrads on YouTube?  Many thanks.</p>&mdash; Dante Scala (@Graniteprof) <a href="https://twitter.com/Graniteprof/status/956997302497021952?ref_src=twsrc%5Etfw">January 26, 2018</a></blockquote>
<script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>
```{r message=FALSE, warning=FALSE, echo = FALSE}
graphs[[3]]
```
## Fourth tweet

<blockquote class="twitter-tweet tw-align-center" data-lang="en"><p lang="en" dir="ltr">Here&#39;s a quick <a href="https://twitter.com/hashtag/rstats?src=hash&amp;ref_src=twsrc%5Etfw">#rstats</a> ggraph snippet for the weekend. I can&#39;t decide which Star Wars 🎬 to watch again. It technically fits in a tweet but doesn&#39;t look great. (Sorry for the earlier broken tweets)<a href="https://t.co/HwF0rp6EIa">https://t.co/HwF0rp6EIa</a> <a href="https://t.co/5ngLsVXdfT">pic.twitter.com/5ngLsVXdfT</a></p>&mdash; Austin Wehrwein (@awhstin) <a href="https://twitter.com/awhstin/status/957003518308700160?ref_src=twsrc%5Etfw">January 26, 2018</a></blockquote>
<script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>
```{r message=FALSE, warning=FALSE, echo = FALSE}
graphs[[4]]
```
A tweet about graphs, how timely `r emo::ji('smiley')`

Poeople doesn't seem angry at all in these replies, they look more like friendly conversations, just as I suspected `r emo::ji('sunglasses')`   

**\#rstats rocks!** `r emo::ji('metal')`
